---
title: "String manipulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following is a very incomplete list of tasks you can accomplish with string manipulation in R. For more, see:

https://rstudio-pubs-static.s3.amazonaws.com/347744_c25378610af54ff5a9b5fc0f0bd9b587.html

```{r, message = FALSE}
library(tidyverse)
```

### Concatenate strings

paste() and paste0() are functions in base R that concatenate vectors (after first converting to character, if necessary).

stringr has a similar function, str_c(), with the advantage that missing values give missing outputs.

These functions are useful for user interfacing.

```{r}
n <- 3
paste("Enter file number ", n, ", please.", sep = "")
paste0("Enter file number ", n, ", please.")
str_c("Enter file number ", n, ", please.")
```

One use case for paste() and str_c() is to build up a file path from a vector of sub-directory names.

```{r}
dir_vector <- c("home", "Documents", "2020_R-workshop", "data")
str_c(dir_vector, collapse = "/")
```

### Extracting substrings

Again, this could be useful for working with filenames. For example, you can strip the extension from a filename. Negative intergers count from the end.

```{r}
filename <- "name.ext"
str_sub(filename, start = 1, end = -5)
str_sub(filename, start = -3, end = -1)
```

### Pattern matching

These stringr functions all take a pattern argument:

    str_detect()
    str_subset()
    str_count()
    str_replace()

Maybe you have a column in a data frame where information about the experiment is embedded in a string.

I'll create one here as an example. Tribble is a convenient function for manually creating a data frame (tibble).

```{r}
exp_info <- tribble(
  ~info,            ~val,
  "control_01_M",   110,
  "control_02_F",   125,
  "case_01_M",      130,
  "case_02_F",      105
)
exp_info

# str_subset() returns strings that match the pattern
exp_info$info %>%
  str_subset("con") # partial matching

# str_detect() returns a logical vector the same length as the original
exp_info$info %>%
  str_detect("con")

# to return the entire row, use dplyr's filter()
exp_info %>%
  filter(str_detect(info, "con")) # filter uses the logical vector to subset the rows
```

### Splitting strings

The info column in the exp_info data frame contains three pieces of information, separated by an underscore:

experiment_number_sex

We can split this string into substrings.

```{r}
string <- exp_info$info[[1]]
string

# By default, str_split() returns a list.
str_split(string, "_")
str(str_split(string, "_"))

# We can simplify the list to a character matrix-
str_split(string, "_", simplify = TRUE)
str(str_split(string, "_", simplify = TRUE))

# or use unlist to produce a character vector-
unlist(str_split(string, "_"))
str(unlist(str_split(string, "_")))

# and access individual elements.
str_split(string, "_", simplify = TRUE)[[1]]
str_split(string, "_", simplify = TRUE)[[2]] # a character
as.integer(str_split(string, "_", simplify = TRUE)[[2]])
```

Use the separate() function in tidyr to split a single character column into multiple columns. It works with either regular expressions or a vector of character positions.The default is to split by any non-alphanumeric value (using the regular expression: sep = "[^[:alnum:]]+"). Use convert = TRUE to convert number from a character to an integer.

```{r}
exp_info %>% 
  separate(info, c("experiment", "number", "sex"), convert = TRUE)
```

### Matching patterns with regular expressions

Regular expressions (regexps) are a very concise way to describe patterns in strings. Regular expressions is a whole topic in itself! Here, we'll touch on the very basics. Also check out:

https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions

https://rstudio-pubs-static.s3.amazonaws.com/347744_c25378610af54ff5a9b5fc0f0bd9b587.html#regular-expressions

The following very helpful tutorial was written for python, but you can apply the concepts to regular expressions in R. It has lots of biologically-relevant examples. I highly recommend taking a look at it when you think you want to learn more about regexps.

https://pythonforbiologists.com/regular-expressions

##### Basic matches

str_view() provides a convenient way to show regular expression matches. I needed to install "htmlwidgets": install.packages("htmlwidgets").

```{r}
# Exact matching
x <- exp_info$info
x
str_view(x, "con")

# A period (.) matches any character.
y <- c("GCACATGGACCAC", "GCACATGGTCCAC", "GCACATGGGCCAC", "GCACATGGCCCAC", "GCACATGGNCCAC")
str_view(y, "GG.CC")

# To match a period, you have to escape its special behavior with a backslash (\).
str_view(filename, "\\.")

# Match a group of characters
str_view(y, "GG(A|T)CC") # alternation, (A|T) matches either A or T
str_view(y, "GG[ATGC]CC") # character group within []
str_view(y, "GG[^N]CC") # [^] match any character except those within group
```

##### Anchors

"Anchor" a regular expression to match the start or end of a string.

Anchoring to the start of a string (^) can be useful in identifying common headers.

```{r}
fasta_lines <- c("> my_DNA_seq", "ATGCCG", "> my_RNA_seq", "AUGCCGAAAAAAA")
str_view(fasta_lines, "^>") # anchor to start
str_view(fasta_lines, "A$") # anchor to end
```

##### Repetition

Control how many times a pattern matches:

    ?: 0 or 1
    +: 1 or more
    *: 0 or more

```{r}
str_view(fasta_lines, "A+$")
```

You can also specify the number of matches precisely:

    {n}: exactly n;
    {n,}: n or more;
    {,m}: at most m;
    {n,m}: between n and m.

Note that these matches are “greedy” by default (they'll match the longest string possible). To make them “lazy” (matching the shortest string possible) put a ? after them.

```{r}
str_view(fasta_lines, "A{5,}")
```

##### Grouping and backreferences

These are more advanced, and you can check them out here.

https://r4ds.had.co.nz/strings.html#grouping-and-backreferences

##### Usage

How do you actually make use of regexps in your code? Regexps can be used with other stringr functions. The example below uses regexps in the function str_detect().

```{r}
# Subset a vector
str_detect(fasta_lines, "A{5,}") # returns a logical vector, can be used for indexing.
fasta_lines[str_detect(fasta_lines, "A{5,}")]

# or a data frame.
exp_info %>% 
  filter(str_detect(info, "M$"))
```

##### Exercises

From https://pythonforbiologists.com/regular-expressions

Write a program that will print only the accession names that satisfy the following criteria – treat each criterion separately:

    contain the number 5
    contain the letter d or e
    contain the letters d and e in that order
    contain the letters d and e in that order with a single letter between them
    contain both the letters d and e in any order
    start with x or y
    start with x or y and end with e
    contain three or more digits in a row
    end with d followed by either a, r or p



