---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Regular expression exercises

From https://pythonforbiologists.com/regular-expressions

Write a program that will print only the accession names that satisfy the following criteria – treat each criterion separately:

    contain the number 5
    contain the letter d or e
    contain the letters d and e in that order
    contain the letters d and e in that order with a single letter between them
    contain both the letters d and e in any order
    start with x or y
    start with x or y and end with e
    contain three or more digits in a row
    end with d followed by either a, r or p


```{r}
accessions <- c('xkn59438', 'yhdck2', 'eihd39d9', 'chdsye847', 'hedle3455', 'xjhd53e', '45da', 'de37dp')

str_view_all(accessions, "5")

str_view_all(accessions, "(d|e)")

str_view(accessions, "d.*e")

str_view_all(accessions, "d.e") # could make this specific to any letter
str_view_all(accessions, "d[A-Za-z]e") # any letter


```